---
title: "Midterm Project"
author: "Camille Okonkwo"
date: "2024-02-26"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(broom)
library(table1)
library(survival)
library(ggsurvfit)
library(survival)
library(lmtest)
library(gt)
library(survMisc)
library(MASS)
```

### Step 0: Data Preparation
```{r step_0}
# loading data
midtermdata = read_csv("data/midtermdata.csv")

# combining categories, creating factor variables, and labelling

# Recode the variables
midtermdata$tx <- factor(
  midtermdata$tx,
  levels = c("0", "1"))

midtermdata$strat2 <- factor(
  midtermdata$strat2,
  levels = c("1", "0"))

midtermdata$sex <- factor(
  midtermdata$sex,
  levels = c("2", "1"))
  
midtermdata$raceth <- factor(
  ifelse(midtermdata$raceth %in% c(3, 4, 5, 6), 6, midtermdata$raceth),
  levels = c("1", "2", "6"))

midtermdata$ivdrug <- factor(
  ifelse(midtermdata$ivdrug %in% c(2, 3), 2, midtermdata$ivdrug), 
  levels = c("2", "1"))

midtermdata$hemophil <- 
  factor(midtermdata$hemophil, 
         levels = c("0", "1"))

midtermdata$karnof <- factor(
  ifelse(midtermdata$karnof %in% c(80, 70), 70, midtermdata$karnof),
  levels = c("100", "90", "70"))
```

# Descriptive statistics
```{r}
# data exploration
summary(midtermdata)

# Summarize Baseline Characteristics
library(table1)

# Define variable labels
table1::label(midtermdata$tx) <- "Treatment"
table1::label(midtermdata$age) <- "Age at Enrollment"
table1::label(midtermdata$cd4) <- "Baseline CD4 Count (cells/mL)"
table1::label(midtermdata$karnof) <- "Karnofsky Performance Scale*"
table1::label(midtermdata$ivdrug) <- "IV Drug Use History*"
table1::label(midtermdata$strat2) <- "CD4 Stratum at Screening"
table1::label(midtermdata$sex) <- "Sex"
table1::label(midtermdata$raceth) <- "Race/Ethnicity*"
table1::label(midtermdata$hemophil) <- "Hemophiliac"
table1::label(midtermdata$priorzdv) <- "Months of prior ZDV use"

# Create a summary table with variable labels
summary_table <- table1(
  ~ age + cd4 + karnof + ivdrug + strat2 + sex + raceth + priorzdv | tx,
  data = midtermdata, 
  footnote = "*Race/Ethnicity, Karnofsky Performance Scale, and IV drug use history variables were recoded due to low patient counts.",
  caption = "Descriptive Statistics of Baseline Characteristics in HIV Clinical Trial Participants")

print(summary_table)
```

is there sufficient evidence that the three-drug regimen has better PFS compared to the two-drug regimen?
```{r PFS_step1}
library(survival)

# PFS K-M table
km_pfs = survfit(Surv(time, censor) ~ tx, data = midtermdata, conf.type = "log-log")

summary(km_pfs)

# log-rank test
log_rank_PFS <- survdiff(Surv(time, censor) ~ tx, data = midtermdata)

# Extract relevant information
chi_squared <- log_rank_PFS$chisq
degrees_of_freedom <- length(log_rank_PFS$n) - 1
p_value <- pchisq(chi_squared, degrees_of_freedom, lower.tail = FALSE)

# Create a custom table
PFS_logrank <- data.frame(
  "Chi-Squared" = chi_squared,
  "Degrees of Freedom" = degrees_of_freedom,
  "P-Value" = p_value
)

# Display the custom table
knitr::kable(PFS_logrank, caption = "Log-Rank Test Results for PFS", digits = 5)

library(ggsurvfit)

# PFS K-M plot
survfit2(Surv(time, censor) ~ tx, data = midtermdata) |> 
  ggsurvfit() +
  add_pvalue(location = "annotation", 
             caption = "Log-rank {p.value}") +
  add_confidence_interval() +
  add_risktable() +
  add_censor_mark() +
  add_legend_title(title = "PFS Curve")
```
#### Our hypotheses:
( where s1(t) is tx = 0 and s2(t) is tx = 1)
* H0 :S1(t) = S2(t), for all t ≤ τ
* Hα :One of the Sk (t) is different for some t ≤ τ

#### Test statistic:
* Q_log-rank = 10.5

#### Degree of freedom: 
* df = 1

#### P-value: 
* Pr(χ21 ≥  10.5) = 0.00117 < 0.05

#### Conclusion: 
We reject H0 at the significance level 0.05. The survival curves for patients in two hormone therapy groups are significantly different, and there is sufficient evidence to conclude that the three-drug regimen has better PFS compared to the two-drug regimen.

```{r pfs_wilx}
# repeat with wilxocon
library(survMisc)

pfs_wilx = ten(survfit(Surv(time, censor) ~ tx, data = midtermdata))
comp(pfs_wilx)
knitr::kable(attributes(pfs_wilx)$lrt[, c(1, 6:8)], "simple", digits = 4)

# Same conclusion with Wilxocon Rank Test
```


are there any other variables significantly associated with PFS?
```{r pfs_step2}
library(survival)
library(lmtest)
library(gt)

# age
age_pfs = coxph(Surv(time, censor) ~ age,
            data = midtermdata,
            ties = "efron")
summary(age_pfs)

# cd4
cd4_pfs = coxph(Surv(time, censor) ~ cd4,
            data = midtermdata,
            ties = "efron")
summary(cd4_pfs)

# hemophil
hemophil_pfs = coxph(Surv(time, censor) ~ hemophil,
            data = midtermdata,
            ties = "efron")
summary(hemophil_pfs)

# ivdrug
ivdrug_pfs = coxph(Surv(time, censor) ~ ivdrug,
            data = midtermdata,
            ties = "efron")
summary(ivdrug_pfs)

# karnof
karnof_pfs = coxph(Surv(time, censor) ~ karnof,
            data = midtermdata,
            ties = "efron")
summary(karnof_pfs)

# priorzdv
priorzdv_pfs = coxph(Surv(time, censor) ~ priorzdv,
            data = midtermdata,
            ties = "efron")
summary(priorzdv_pfs)

# raceth
raceth_pfs = coxph(Surv(time, censor) ~ raceth,
            data = midtermdata,
            ties = "efron")
summary(raceth_pfs)

# sex
sex_pfs = coxph(Surv(time, censor) ~ sex,
            data = midtermdata,
            ties = "efron")
summary(sex_pfs)

# strat2
strat2_pfs = coxph(Surv(time, censor) ~ strat2,
            data = midtermdata,
            ties = "efron")
summary(strat2_pfs)

# tx
tx_pfs = coxph(Surv(time, censor) ~ tx,
            data = midtermdata,
            ties = "efron")
summary(tx_pfs)

# multivariate Cox model
multivar_pfs = coxph(Surv(time, censor) ~ tx + sex + age + karnof + cd4 + strat2,
            data = midtermdata,
            ties = "efron")

summary(multivar_pfs)
```

reducing multivariable model using log partial likelihood ratio test
```{r}
# LRT using lmtest package
library(lmtest)

r_model <- coxph(Surv(time, censor) ~ tx + sex + age + karnof + cd4, data = midtermdata, ties = "efron")

f_model <- coxph(Surv(time, censor) ~ tx + sex + age + karnof + cd4 + strat2 + raceth + ivdrug + hemophil + priorzdv, data = midtermdata, ties = "efron")

lrtest(f_model, r_model)
# since there is insufficient evidence to conclude the reduced model is better than the full, I will try another method.

# selection tests
library(MASS)
stepwise_cox_model <- stepAIC(f_model, direction = "both")
summary(stepwise_cox_model)
lrtest(f_model, stepwise_cox_model)
# There is insufficient evidence to conclude that the stepwise model is better than the full model. 

lrtest(stepwise_cox_model, r_model)
# Since the step wise model and reduced model are not significantly better than the full model, for the sake of parsimony I will choose the reduced model. 
```

Are there any significant effect modifiers for the relation between the treatment regimens and PFS?
```{r}
# test interactions (pfs 2)
multivar_pfs2 = coxph(Surv(time, censor) ~ tx + sex + age + karnof + cd4 + cd4*karnof, 
                      data = midtermdata, 
                      ties = "efron")
summary(multivar_pfs2)

lrtest(r_model, multivar_pfs2)
```
There are no significant effect modifiers for the relation between the treatment regimens and PFS.

```{r}
# final model
final_pfs <- coxph(Surv(time, censor) ~ tx + sex + age + karnof + cd4, data = midtermdata, ties = "efron")
```


Is there sufficient evidence that the three-drug regimen has better OS compared to the two-drug regimen?
```{r OS_step1}
# OS K-M table
km_os = survfit(Surv(time_d, censor_d) ~ tx, data = midtermdata, conf.type = "log-log")

summary(km_os)

# log-rank test OS
log_rank_OS <- survdiff(Surv(time_d, censor_d) ~ tx, data = midtermdata)

library(ggsurvfit)

# K-M plot
survfit2(Surv(time_d, censor_d) ~ tx, data = midtermdata) |> 
  ggsurvfit() +
  add_pvalue(location = "annotation", 
             caption = "Log-rank {p.value}") +
  add_risktable() +
  add_confidence_interval() +
  add_censor_mark() +
  add_legend_title(title = "OS Curve")
```
#### Our hypotheses:
( where s1(t) is tx = 0 and s2(t) is tx = 1)
* H0 :S1(t) = S2(t), for all t ≤ τ
* Hα :One of the Sk (t) is different for some t ≤ τ

#### Test statistic:
* Q_log-rank = 4.1

#### Degree of freedom: 
* df = 1

#### P-value: 
* Pr(χ21 ≥  4.1) = 0.0438 < 0.05

#### Conclusion: 
We reject H0 at the significance level 0.05. The survival curves for patients in two hormone therapy groups are significantly different, and there is sufficient evidence to conclude that the three-drug regimen has better OS compared to the two-drug regimen.

are there any other variables significantly associated with OS?
```{r OS_step2}
# univariate Cox models
# age
age_os = coxph(Surv(time_d, censor_d) ~ age,
            data = midtermdata,
            ties = "efron")
summary(age_os)

# cd4
cd4_os = coxph(Surv(time_d, censor_d) ~ cd4,
            data = midtermdata,
            ties = "efron")
summary(cd4_os)

# hemophil
hemophil_os = coxph(Surv(time_d, censor_d) ~ hemophil,
            data = midtermdata,
            ties = "efron")
summary(hemophil_os)

# ivdrug
ivdrug_os = coxph(Surv(time_d, censor_d) ~ ivdrug,
            data = midtermdata,
            ties = "efron")
summary(ivdrug_os)

# karnof
karnof_os = coxph(Surv(time_d, censor_d) ~ karnof,
            data = midtermdata,
            ties = "efron")
summary(karnof_os)

# priorzdv
priorzdv_os = coxph(Surv(time_d, censor_d) ~ priorzdv,
            data = midtermdata,
            ties = "efron")
summary(priorzdv_os)

# raceth
raceth_os = coxph(Surv(time_d, censor_d) ~ raceth,
            data = midtermdata,
            ties = "efron")
summary(raceth_os)

# sex
sex_os = coxph(Surv(time_d, censor_d) ~ sex,
            data = midtermdata,
            ties = "efron")
summary(sex_os)

# strat2
strat2_os = coxph(Surv(time_d, censor_d) ~ strat2,
            data = midtermdata,
            ties = "efron")
summary(strat2_os)

# tx
tx_os = coxph(Surv(time_d, censor_d) ~ tx,
            data = midtermdata,
            ties = "efron")
summary(tx_os)


# multivariate Cox model
multivar_os = coxph(Surv(time_d, censor_d) ~ tx + sex + age + karnof + cd4 + strat2,
            data = midtermdata,
            ties = "efron")

summary(multivar_os)
```

reducing multivariable model using log partial likelihood ratio test
```{r}
full_model <- coxph(Surv(time_d, censor_d) ~ tx + age + sex + cd4 + strat2 + karnof + raceth + ivdrug + hemophil + priorzdv, data = midtermdata, ties = "efron")

reduced_model <- coxph(Surv(time_d, censor_d) ~ tx + sex + age + karnof, data = midtermdata, ties = "efron")
summary(reduced_model)

# LRT using lmtest package
lrtest(full_model, reduced_model)

# since there is insufficient evidence to conclude the reduced model is better than the full, I will try another method.

# selection tests
library(MASS)
stepwise_cox_model2 <- stepAIC(full_model, direction = "both")
summary(stepwise_cox_model2)

lrtest(full_model, stepwise_cox_model2)

# There is insufficient evidence to conclude that the stepwise model is better than the full model. 

# is the stepwise model better than the reduced model?
lrtest(reduced_model, stepwise_cox_model2)
```
For the sake of parsimony, I will choose my initial reduced model over the stepwise model.

Are there any significant effect modifiers for the relation between the treatment regimens and OS?
```{r}
# test interactions (OS 2)
multivar_os2 = coxph(Surv(time_d, censor_d) ~ tx + sex + age + karnof + karnof*tx, 
                      data = midtermdata, 
                      ties = "efron")
summary(multivar_os2)
summary(multivar_os)
```
There are no significant effect modifiers for the relation between the treatment regimens and OS.

```{r}
#final model
final_os <- coxph(Surv(time_d, censor_d) ~ tx + sex + age + karnof, data = midtermdata, ties = "efron")
summary(reduced_model)
```

